---
title: "Covid project"
author: "Kristian Riis Hansen"
date: '2022-09-06'
output: html_document
---
Libraries:
```{r}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(lubridate)

```

working directory
```{r}
setwd("C:/Users/Krist/Desktop/PiB/Variants and mutation rate in SARS-cov2")
getwd()

```


Loading data: 
```{r}


tidy_300K <- readRDS("metadata_snpeff_tidy_300K_downsampled.RDS")
mutations_tidy_300K <- readRDS("mutations_snpeff_annotated_tidy_300K_downsampled.rds")

```



#Are virus mutation rates correlated with location/different pangolin lineages?
#GC content

#Are there single mutations that indicate increased spread in populations. (more virulent or something)


```{r}

count(tidy_300K$country)
?count

table(tidy_300K$country)
countries_over_1k <- tidy_300K %>% group_by(country) %>% mutate(Count=n())

countries_over_1k

```



#new dataframe with count over 1k.
```{r}


filterdf <- countries_over_1k %>% filter(Count >1000)

```



#Shows countries with over 1000 samples
```{r}
table(filterdf$country)

View(filterdf)



small_filt <- filterdf %>% select (country,id, `Virus name`, count_N, count_S) %>%
  mutate(S_over_N = count_S/count_N)

```



  
#Removing NaN rows:

```{r}
small_filt <- na.omit(small_filt)

small_filt <- small_filt[is.infinite(rowSums(small_filt)),]


```



#There are still rows with "inf"

```{r}

small_filt %>% 
  ggplot(aes(country, S_over_N)) +
  geom_boxplot() +
  geom_jitter(width = 0.15) +
  theme(axis.text.x = element_text(angle = 45))


?geom_boxplot()


small_filt %>% 
  ggplot(aes(country, S_over_N)) + 
  geom_boxplot() + 
  geom_jitter(width = 0.2) + 
  theme(axis.text.x = element_text(angle = 45))

```








Merging gene coloumn from tidy_300k to mutation dataset:


we want to detect the selective sweep
we found a method for that in one of the articles - so we count the mutations for each positions and we got that around position 22.500 there are significantly more datapoints so we want to analyse that specific region and check the certain gene which might be responsible for the sweep.

```{r}
left <- tidy_300K

right <- mutations_tidy_300K %>% select(1,5)

merged_df <- merge(left, right)

```

```{r}
left_join(left,right, by =c("id"="id", "position"="position"))

```




We want to merge the datasets so we have dates, country, GC-content.



```{r}


test_mutation_df <- mutations_tidy_300K[,]
  
test_tidy_df <- (tidy_300K[,c(1,10,13,31)])

test_mutation_df <- test_mutation_df %>% group_by(id)


test_merged_df <- left_join(test_tidy_df, test_mutation_df, by =c("id"="id"))


test_merged_df <- test_merged_df %>% group_by(position) %>% 
  mutate(count = n())


```




We now look at a specific country and gene and graph some things.

```{r}

dk_merged <- filter(test_merged_df, country == "Denmark")


fra_merged <- filter(test_merged_df, country == "France")

```

Converting date column into numeric. Works for denmark because it has days included. but the ones in france that already have month only, will get NA.

```{r}

Date = format(as.Date(dk_merged$`Collection date`), "%Y - %m")


Date = format(as.Date(fra_merged$`Collection date`), "%Y-%m")


#Trying using substring
fra_merged <- mutate(fra_merged, Date = format(substring(fra_merged$`Collection date`,0,7)))
#Works like a charm.

dk_merged <- dk_merged %>% mutate(Date = Date)


```




Looking at GC-content over time in Denmark.

```{r}

ggplot(data=dk_merged, aes(x=Date, y=dk_merged$"GC-Content")) + 
         geom_boxplot() +
        theme(axis.text.x=element_text(angle=90)) + ylab("GC content")
      



```
GC-content over time in France.

Because france has samples that are labeled on only the years, we will filter these out.

Using grepl to remove the "years only" rows.
```{r}

remove <- c("2020,2021,2022")

fra_merged <- fra_merged %>% grepl(remove, fra_merged$Date)






```


Plotting GC content for france over time.
```{r}

ggplot(data=fra_merged, aes(x=Date, y=fra_merged$"GC-Content")) + 
         geom_boxplot() +
        theme(axis.text.x=element_text(angle=90)) + ylab("GC content")
      



```

Filter danish dataset for only one year and spike gene to see mutation positions.


```{r}

dk_2020 <- dk_merged %>% filter(str_detect(Date, "^2020"))




dk_2020 %>% group_by(position) %>%
  mutate(count = n()) %>%
  filter(gene == "S") %>%
  ggplot(dk_2020, mapping = aes(x=position, y=count)) + geom_point() +
  ggtitle("Spike gene mutations during 2020 in Denmark")



```

Same for france:

```{r}

fra_merged %>% filter(str_detect(Date,"^2020")) %>% 
  group_by(position) %>% 
  mutate(count = n()) %>% 
  filter(gene == "S") %>% 
  ggplot(fra_merged, mapping = aes(x=position, y=count)) + geom_point() + 
  ggtitle("Spike gene mutations during 2020 in France")



```


```{r}

unique(tidy_300K$pangolin_lineage)


```



Mutations over time for the "5 main lineages"


```{r}



tidy_300K_test <- mutate(tidy_300K, Date = format(substring(tidy_300K$`Collection date`,0,7)))



#Here we use lubridate to make the collection date column into an actual DATE.
tidy_300K_test$`Collection date` <- ymd(tidy_300K_test$`Collection date`) 

  



unique(tidy_300K_test$Variant)



  

tidy_300K_test <- tidy_300K_fra %>% mutate(mutations = (count_N + count_S))

interesting_lineages <- c("B.1.1.7", "P.1", "B.1","B.1.351")

tidy_300K_test <- tidy_300K_test %>% mutate(pango_lineage = ifelse(pangolin_lineage %in% interesting_lineages, pangolin_lineage, "Other lineages"))





tidy_300K_test %>%
  group_by(id, mutations) %>%
  filter(str_detect(Variant, "^VOC")) %>% #This is saying the Variants only starting with "VOC"
  ggplot(tidy_300K_test, mapping = aes(x=`Collection date`,y=mutations, color=Variant)) +
  geom_point() +
  theme(axis.text.x=element_text(angle=90)) +
  theme(legend.title = element_text(size = 3),
         legend.text = element_text(size = 3))+
  ggtitle("Mutations over time for the common lineages")


#Try and remove 2020 2021


?nchar

```


GC content of the 4-5 lineages of interest over time.

```{r}



tidy_300K_test %>%
  group_by(id) %>%
  filter(str_detect(Variant, "^VOC")) %>% #This is saying the Variants only starting with "VOC" 
  ggplot(tidy_300K_test, mapping = aes(x=`Collection date`,y=`GC-Content`, color=Variant)) +
  geom_point(size = 2) +
  geom_smooth(aes(color=Variant), method = "lm", se = F) +
  theme(axis.text.x=element_text(angle=90)) +
  theme(legend.title = element_text(size = 4),
         legend.text = element_text(size = 4)) +
  ggtitle("GC-Content over time for the common lineages") +
  NULL




```



Looking at France only:
```{r}

tidy_300K_test %>% filter(country == "France") %>% 
  filter(str_detect(Variant, "^VOC")) %>% #This is saying the Variants only starting with "VOC" 
  group_by(id) %>%
  ggplot(tidy_300K_test, mapping = aes(x=`Collection date`,y=`GC-Content`, color=Variant)) +
  geom_point() +
  geom_boxplot() +
  theme(axis.text.x=element_text(angle=90)) +
  theme(legend.title = element_text(size = 3),
         legend.text = element_text(size = 3))+
  ggtitle("GC-Content over time for the common lineages in France")+
  facet_wrap( ~ Variant)
  



```



dn/ds for each lineage


```{r}

tidy_300K_test <- tidy_300K_test %>% mutate(DnDs = (count_N/count_S))


tidy_300K_test %>% filter(country == "France") %>% 
  group_by(id) %>%
  ggplot(tidy_300K_test, mapping = aes(x=Date,y=DnDs, color=pango_lineage)) +
  geom_point() +
  theme(axis.text.x=element_text(angle=90)) +
  ggtitle("DnDs ratio for the common lineages in France")+
  facet_wrap( ~ pango_lineage)
  



```


We want to find out whether the mutations are often G/C or otherwise. To do so we look at the reference genome: Wuhan1 and by using a selfwritten function, we use the mutation dataset to see where aswell as which mutation. So 1. we can see which mutations are most prone and also in which regions. We do this because we hypothesize that there are more mutations in the region for immunerecognision associated genes.




```{r}

x <- read_tsv("NC_045512.2.fasta")
reference_genome <- strsplit(x$`>NC_045512.2 Severe acute respiratory syndrome coronavirus 2 isolate Wuhan-Hu-1, complete genome`[1], "") %>% unlist()
rm(x)
table(reference_genome) %>% knitr::kable()
 

ref_genome <- tibble(base = reference_genome) %>% mutate(position=row_number())




#do it with the frequencies of A C G T?
```



```{r}



bases <- c("T","G","C","A")

dk_merged <- test_merged_df %>% filter(country == "Denmark") %>% filter(variant_base == bases)

#now only variant_base = A T C or G


#Join the wuhan genome to the dk_merged:
  

left_dk <- dk_merged

right_dk <- ref_genome 


dk_merged_with_wuhan <- left_join(left_dk, right_dk, by = "position")


```



```{r}

  
  #Just for removing a column:
#dk_merged_with_wuhan <- subset(dk_merged_with_wuhan, select = -newCol)

#Adding Variants to the dk_merged_with_wuhan with merged:

right <- tidy_300K %>% select(Variant,id)
left <- dk_merged_with_wuhan

dk_merged_with_wuhan <- left_join(left,right, by = "id")

dk_merged_with_wuhan$`Collection date` <- ymd(dk_merged_with_wuhan$`Collection date`)
#Now we go into pyhton and load the "dk_merged_wuhan.csv"

write.csv(dk_merged_with_wuhan,"dk_merged_wuhan.csv")


dk_wuhan_statcount <- read.csv("dk_wuhan_statcount.csv")
dk_wuhan_statcount_with_Variant <- read.csv("dk_wuhan_statcount_with_Variant.csv")

```


Wrangling dataset with base change counts:

We want to produce a plot showing the occurences of changes for each position. Which variant base is most likely to shift to. We do this to get an understanding and insight into the possibility that different regions of the genome might be under different selective pressures. Coupled with the GC-content.
```{r}

#Only VOC, not VOI:

dk_wuhan_statcount_with_Variant <- dk_wuhan_statcount_with_Variant %>% filter(str_detect(Variant, "^VOC"))

dk_merged_with_wuhan <- dk_merged_with_wuhan %>% filter(str_detect(Variant, "^VOC"))

#Shortening the VOC variable names:
unique(dk_wuhan_statcount_with_Variant$Variant)

library(stringr)

dk_wuhan_statcount_with_Variant <- dk_wuhan_statcount_with_Variant %>% mutate(variant = str_extract(dk_wuhan_statcount_with_Variant$Variant, "[^(]+"))
#thank fucking god this works.



dk_wuhan_statcount_with_Variant %>% group_by(position,variant) %>% 
  ggplot(aes(x=position, y=count, fill = variant_base))+
  geom_bar(stat="identity")




dk_wuhan_statcount_with_Variant %>% group_by(position) %>% 
  ggplot((aes(x=position, y=count, color=variant_base)))+
  xlim(22300,24000)+
  ylim(0,10)+
  geom_point() +
  ggtitle("counts of limited mutations for each VOC in range of 22300-24000")+
  facet_wrap(~ variant)


```


Frequency table:

```{r}


dk_wuhan_statcount_with_Variant$Pairs <- paste(dk_wuhan_statcount_with_Variant$base, dk_wuhan_statcount_with_Variant$variant_base)

    

frequencies_with_Variant <- dk_wuhan_statcount_with_Variant %>% group_by(Pairs,variant) %>% summarise(sum(count)/184171,sum(count))



sum(frequencies_with_Variant$`sum(count)/184171`)
frequencies_with_Variant




```

Frequency plots: one for the general tendency and then for some specific regions:


```{r}

#Count plot for each change over the whole genome:
ggplot(frequencies_with_Variant, aes(x=Pairs,y=frequencies_with_Variant$`sum(count)`,color=variant))+
         geom_point()+
  theme(axis.text.x=element_text(angle=90)) +
  ggtitle("Counts of changes over the whole genome")



```

```{r}


#Count plot for each change in the 22500- 24000 region:


frequencies_in_range <- dk_wuhan_statcount_with_Variant %>% filter(between(position,22500,24000)) %>% group_by(Pairs,variant) %>% summarise(sum(count)/184171,sum(count))


ggplot(frequencies_in_range,aes(x=Pairs,y=frequencies_in_range$`sum(count)`, color=variant))+
  geom_point()+theme(axis.text.x=element_text(angle=90)) +
  ylim(0,10)+
  geom_jitter()+
  ggtitle("Counts of small changes in range 22500 - 24000 for all VOC")
  

```



Short variant names for dk_merged_with_wuhan so we can make a GC graph for all VOCs
```{r}

dk_merged_with_wuhan <- dk_merged_with_wuhan %>% mutate(variant = str_extract(dk_merged_with_wuhan$Variant, "[^(]+"))

dk_merged_with_wuhan %>% filter(variant =)
ggplot(dk_merged_with_wuhan, aes(x=dk_merged_with_wuhan$`Collection date`,y=dk_merged_with_wuhan$`GC-Content`, color=variant)) +
  geom_point() +
  geom_smooth(aes(color=variant), method = "lm", se = F) +
  ggtitle("GC-content of Variants of concern in Denmark")


```




Adding aa_change to statcount aswell:
```{r}

#Going into python environment and adding the column aswell:

```



L452R over timeline:

```{r}


dk_wuhan_statcount_with_Variant_Collectiondate <- read.csv("dk_wuhan_statcount_with_Variant_and_collection_date.csv")

#adding variant short:
#Filtering only for VOC:
dk_wuhan_statcount_with_Variant_Collectiondate <- dk_wuhan_statcount_with_Variant_Collectiondate %>% filter(str_detect(Variant, "^VOC"))

#making variant column
dk_wuhan_statcount_with_Variant_Collectiondate <- dk_wuhan_statcount_with_Variant_Collectiondate %>% mutate(variant = str_extract(dk_wuhan_statcount_with_Variant_Collectiondate$Variant, "[^(]+"))

#lubridate the date
dk_wuhan_statcount_with_Variant_Collectiondate$Collection.date <- ymd(dk_wuhan_statcount_with_Variant_Collectiondate$Collection.date)



dk_wuhan_statcount_with_Variant_Collectiondate %>% filter(aa_change == "L452R") %>%   ggplot(aes(x=Collection.date,y = count, color=variant))+
  geom_point() +
  ggtitle("Number of mutations that cause the L452R change")




```


Where is the spike gene:

We expect to see 






```{r}

#NTD
21563+13

21563+304

#RBD
21563+319
21563+541


unique(dk_merged_with_wuhan$effect)

syn <- "synonymous_variant"

#dk_merged_with_wuhan %>% mutate(Type = ifelse(effect %in% syn, effect, "Non-syn"))

test <- dk_merged_with_wuhan %>% group_by(variant, position) %>% summarise(n=n()) %>% mutate(counts = n)

dk_merged_with_wuhan %>% summarise(n=n())


dk_merged_with_wuhan %>% mutate(Type = ifelse(effect %in% syn, effect, "Non-syn")) %>% filter(gene=="S") %>% filter(between(position, 21500,22150)) %>% group_by(variant, position, Type) %>% summarise(n=n()) %>% filter(n <= 5) %>% 
ggplot()+
  geom_point(aes(x=position, y=n,color = Type))+
  facet_wrap(~ variant)+
    ggtitle("Low frequency mutations on the NTD & RBD regions")


```





A plot showing all the positions without a mutation: for all different VOC. they show the same pattern?
```{r}




```






test hvordan group by virker: DET GØR DET SÅDAN HER BRO
```{r}


dk_wuhan_statcount_with_Variant_Collectiondate %>% group_by(position,variant,aa_change,base,variant_base) %>% summarise(k=n()) %>% 
  ggplot((aes(x=position, y=k, color=variant_base)))+
  xlim(22300,24000)+
  ylim(0,10)+
  geom_point() +
  ggtitle("counts of limited mutations for each VOC in range of 22300-24000")+
  facet_wrap(~ variant)


```

