---
title: "Covid project"
author: "Kristian Riis Hansen"
date: '2022-09-06'
output: html_document
---
Libraries:
```{r}
library(dplyr)
library(tidyverse)
library(ggplot2)


```

working directory
```{r}
setwd("C:/Users/Krist/Desktop/PiB/Variants and mutation rate in SARS-cov2")
getwd()

```


Loading data: 
```{r}


tidy_300K <- readRDS("metadata_snpeff_tidy_300K_downsampled.RDS")
mutations_tidy_300K <- readRDS("mutations_snpeff_annotated_tidy_300K_downsampled.rds")

```



#Are virus mutation rates correlated with location/different pangolin lineages?
#GC content

#Are there single mutations that indicate increased spread in populations. (more virulent or something)


```{r}

count(tidy_300K$country)
?count

table(tidy_300K$country)
countries_over_1k <- tidy_300K %>% group_by(country) %>% mutate(Count=n())

countries_over_1k

```



#new dataframe with count over 1k.
```{r}


filterdf <- countries_over_1k %>% filter(Count >1000)

```



#Shows countries with over 1000 samples
```{r}
table(filterdf$country)

View(filterdf)



small_filt <- filterdf %>% select (country,id, `Virus name`, count_N, count_S) %>%
  mutate(S_over_N = count_S/count_N)

```



  
#Removing NaN rows:

```{r}
small_filt <- na.omit(small_filt)

small_filt <- small_filt[is.infinite(rowSums(small_filt)),]


```



#There are still rows with "inf"

```{r}

small_filt %>% 
  ggplot(aes(country, S_over_N)) +
  geom_boxplot() +
  geom_jitter(width = 0.15) +
  theme(axis.text.x = element_text(angle = 45))


?geom_boxplot()


small_filt %>% 
  ggplot(aes(country, S_over_N)) + 
  geom_boxplot() + 
  geom_jitter(width = 0.2) + 
  theme(axis.text.x = element_text(angle = 45))

```



Mutation dataset with only position > 20000:

```{r}


Above_20k_position <- mutations_tidy_300K %>% filter(position>20000)

unique(Above_20k_position$gene)
#still 19 genes showing in the last 10k positions.

```





A plot looking at a section or all of the mutation positions --> with counts of these point mutations as Y.




#Does not really work with histogram because of bins etc.
```{r}

mutations_tidy_300K %>%
  ggplot() + 
  geom_histogram(aes(position), binwidth = 1, fill = mutations_tidy_300K$gene)
  

  
ggplot(Above_20k_position, aes(x=position), fill = gene) +
         geom_histogram(binwidth = 1)


```


Creating same plot but with geom_point, creating a coloumn for count.
```{r}



Above_20k_position <- Above_20k_position %>% group_by(position) %>% 
  mutate(count = n())






Above_20k_position %>% group_by(count,position)

  ggplot(Above_20k_position, aes(x=position, y=count, colour = gene)) + geom_point()

```

Filter for only spike gene:
```{r}


Spike_gene <- filter(Above_20k_position, gene == "S")

ggplot(Spike_gene, aes(x=position, y=count)) + geom_point()


```

Further removing synonymous variants:
```{r}
unique(Spike_gene$effect)

Spike_gene_noSyn <- filter(Spike_gene, effect != "synonymous_variant")


#Filtering for only high and modifier impact
unique(Spike_gene_noSyn$impact)

Spike_gene_noSyn <- filter(Spike_gene_noSyn, impact == "HIGH"|impact == "MODIFIER")




```
Plotting the Spike_gene_filtered

```{r}

ggplot(Spike_gene_noSyn, aes(x=position, y=count)) + geom_point()


```

Merging gene coloumn from tidy_300k to mutation dataset:


we want to detect the selective sweep
we found a method for that in one of the articles - so we count the mutations for each positions and we got that around position 22.500 there are significantly more datapoints so we want to analyse that specific region and check the certain gene which might be responsible for the sweep.

```{r}
left <- tidy_300K

right <- mutations_tidy_300K %>% select(1,5)

merged_df <- merge(left, right)

```

```{r}
left_join(left,right, by =c("id"="id", "position"="position"))

```






